---
title: "Voting Rights & the Supreme Court"
author: "Mallory Bertone"
description: "PS270 Final Project"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(knitr)
```

## Introduction

The Voting Rights Act of 1965 was signed into law by President Lyndon B. Johnson, with the aim of providing direct federal intervention for African Americans to be able to vote, as well as banned tactics that had long been used to keep them from voting. This was a huge expansion of voting rights for Americans, as it aimed at stopping voter suppression that had taken place since the U.S. had been founded. Since 1969, 14 of the 18 justices appointed have been selected by Republican presidents, making them likely to follow the opinions of their appointees, some of which may include limiting access to voting. Currently, the Supreme Court has a 6-3 divide, with the majority leaning conservative. 

**Has the U.S. Supreme Court expanded or limited access to voting since the passing of the Voting Rights Act of 1965?**

## Data

The data used for this analysis comes from the Supreme Court Database from Washington University Law School in St. Louis, which contains over 200 pieces of information from each case the court has heard between the 1791 and 2024 terms. In this case, this data specifically comes from the MODERN database, which includes information from the years 1945 to 2024. 

Our analysis is a cross-sectional study, since this was an observational study that involved looking at data from a population, which in this case is U.S. Supreme Court cases, over a specific period of time. The explanatory variable is the court era, in which each era is decided by the judge that is the Chief Justice. Since 1965, the court eras have included the Warren court, the Burger court, the Rehnquist court, and the Roberts court, which is the era the Supreme Court is currently in. From this, the outcome variable is the outcome of the case, on which we will base whether or not voting access was expanded or restricted. This will be measured by the `decisionDirection` variable, which notes whether the court voted conservatively, liberally, or their decision was unspecifiable. 

When answering this question, the null hypothesis will be that the Supreme Court has restricted voting rights since 1965, since the judges tend to be more conservative than liberal. The alternative hypothesis will be that the Supreme Court has expanded voting rights. If the courts decisions trend conservatively, the hypothesis will be supported. If the court trends liberally or the decision was unspecifiable, the hypothesis will not be supported. 

## Data Cleaning 

```{r loan and clean data, include=FALSE}
cases <- read_csv("SCDB_2025_01_caseCentered_Vote.csv")
cases <- cases |> mutate(year = term) |> 
  select(year, caseId, dateDecision, decisionType, naturalCourt, chief, docket, caseName, 
         caseDisposition, caseDispositionUnusual, partyWinning, voteUnclear, issue, issueArea, 
         decisionType, decisionDirection, authorityDecision1, authorityDecision2, lawType, lawSupp, 
         splitVote, majVotes, minVotes)
voting_list <- c(20010, 20020, 20090)
cases <- cases |> filter(year > 1965, issue == "20010" | issue == "20020" | issue == "20090") 
cases <- cases |> distinct(caseName, .keep_all = TRUE)
cases <- cases |> drop_na(decisionDirection) |> 
  filter(decisionDirection == 1 | decisionDirection == 2) |> 
  mutate(ideology = if_else(decisionDirection == 1, "Conservative", "Liberal"), 
         chief = factor(chief, levels = c("Warren", "Burger", "Rehnquist", "Roberts")))
cases <- cases |> mutate(ideologyBinary = if_else(decisionDirection == 1, 1, 0))
```

```{r display data, echo=FALSE}
head(cases) |> select(year, dateDecision, caseName, chief, decisionDirection, ideology) |> 
  knitr::kable()
```

## Data Visualization

```{r visualization, echo=FALSE}
plot_data <- cases |>
  filter(!is.na(ideologyBinary)) |>
  mutate(ideology_label = factor(ideologyBinary, levels = c(0, 1), 
                                 labels = c("Liberal", "Conservative")))
plot_data |> ggplot(aes(x = ideology_label, fill = ideology_label)) + 
  geom_bar() + theme_minimal() +
  labs(title = "Decision Direction Count Based on Court Rule", 
       x = "Ideology", y = "Count of Cases") +
  facet_wrap(~chief, ncol = 2) + 
  scale_fill_manual(values = c("Conservative" = "red3", "Liberal" = "royalblue")) + 
  theme(legend.position = "none",
    strip.text = element_text(size = 12, face = "bold"))
```

Each bar graph represents a different court era since 1965, and compares the number of Conservative and Liberal decisions made in cases about voting for each era. This shows that, overall, the Warren and Burger courts have expanded voting rights more than they restricted them, since their decisions trend in a Liberal direction. The Rehnquist and Roberts courts, overall, voted to restrict voting rights more than expand them, overall, but their Conservative decisions aren't much higher than their Liberal ones. This shows that the Supreme Court has expanded voting rights, as they've made more Liberal decisions than Conservative ones since the passage of the Voting Rights Act of 1965. 

## Regression Analysis

```{r analysis, echo=FALSE}
cases$issue <- factor(cases$issue)
fit <- lm(ideologyBinary ~ issue + chief, data = cases)
cleaned_fit <- tidy(fit)
cleaned_fit |> knitr::kable(digits = 3)
```

### Regression Results 

The `issue` column in the cases dataset contains three unique variables. Issue 20010 is related to any case that falls under voting. Issue 20020 deals specifically with the Voting Rights Act of 1965 and any proposed amendments to it. Issue 20090 deals with reapportionment. 

The model predicts that for cases regarding 20010 during the Warren court have a 0.173 probability of having a Liberal ruling. Holding the era constant, we can see that cases regarding issue 20020 are 14.9 percentage points less likely to have a Liberal ruling than issue 20010, and cases regarding issue 20090 are 8.4 percentage points more likely to hold a Liberal ruling. Compared to the Warren court, the Burger court is 29.3 percentage points more likely, the Rehnquist court is 36.8 percentage points more likely, and the Roberts court is 36.0 percentage points more likely to make a Liberal ruling on a case. 

For the purpose of this analysis, $\alpha = 0.05$. Issue is not a strong predictor of which direction a case will go, because the p-values, which are 0.139 and 0.396, respectively, are not statistically significant, as they are all about 0.05. This means that issue type does not have a strong effect on the ruling direction of a case. The court eras show a different result. The Burger court has a p-value of 0.015, the Rehnquist court has a p-value of 0.003, and the Roberts court has a p-value of 0.008, all of which are less than 0.05, making them statistically significant. Since they are statistically significant, court era does have a strong effect on the probability of a Liberal ruling. 
From this analysis, we can see that Chief Justice is a stronger predictor of which direction a case decision will go in than issue type is. 

While the data visualization shows that more recent courts have made slightly more Conservative decisions, when we control for legal issues using a regression model, we find a positive shift in the probability of Liberal rulings during the Rehnquist and Roberts eras compared to the Warren era. This shows that the Supreme Court has overall expanded voting rights, even across different challenges and ideas regarding the Voting Rights Act. 

## Conclusion

### Code

```{r print-code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```

